---
title:  |
    | CWP1
    | September 1 2021
author: |
    | Dale Morrell
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  pdf_document:
    number_sections: yes
documentclass: article
classoption: a4paper
fontsize: 11pt
header-includes:
- \renewcommand{\thesection}{Q\arabic{section}}
- \renewcommand{\thesubsection}{Part \alph{subsection}.}
- \usepackage{fullpage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load \& explore the data

```{r loadPackages, message=FALSE, results="hide", echo=FALSE}
requiredPackages <- c("knitr",  "tidyverse", "readr")
packages_to_install <- requiredPackages[
                        !(requiredPackages %in% installed.packages()[,1])]

# Uncomment this line to install the packages
#if(length(packages_to_install)>0 ) install.packages(packages_to_install, 
#                                                    dependencies=TRUE)

# Load
sapply(requiredPackages, library, character.only=TRUE)
 
```

## Dataset
```{r loadData}
library(tidyverse)
# Read in the CSV file
quakes_all <- read_csv("CWPquakes.csv") 
```

## Take a sample of data
```{r}
set.seed(17981807) # Specific seed for testing purposes
quakes <- sample_n(quakes_all, size = 1000)
print(top_n(quakes, 5)) # Display the top 5 rows
```

## Shallow earthquakes with large magnitude

```{r}
shallowquake <- dplyr::filter(quakes, depth <= 40) 

# Utilizing dply Filter Command

top_n(shallowquake, 5) # Display top 5

shallow2 <- dplyr::arrange(shallowquake, desc(shallowquake$magnitude)) 

# Use depth filter to dplyr::Arrange by Magnitude.

top_n(shallow2, 5) # Display top 5

Shallowslice <- dplyr::slice(shallow2, 1:5) 

# Slice command, grab 5 mag + shallow depth

top_n(Shallowslice, 5) # Display top 5
```

# Earthquake Magnitude

This dataset consists of the magnitude of 1000 earthquakes, randomly extracted 
from a larger geological dataset detailing earthquakes over the period of 
a year within New Zealand from the 8th of March, 2020 to the 7th of March, 2021.
The dataset contains a mean(quakes\$magnitude) of Magnitude 2.131 (3 d.p),
with a min(quakes\$magnitude) and max(quakes\$magnitude) of Magnitudes 0.058 
and 5.177 (3 d.p) respectively. Furthermore, the histogram reveals that the 
data appears to be normally distributed, with a slight right skew. 
This can be further identified through the Box plot, revealing a right skew 
with two notable outlying earthquakes above a magnitude of 5.0. The lower 25% 
and upper 75% of the data, as revealed through quantile(quakes\$magnitude), 
lies at magnitudes of 1.622 and 2.487 respectively, and the overall dataset
has a magnitude range(quakes\$magnitude) of 0.058 to 5.177.

```{r mag_describe}
# Summary of the dataset

summary(quakes$magnitude)
mean(quakes$magnitude)
min(quakes$magnitude)
max(quakes$magnitude)
quantile(quakes$magnitude)
range(quakes$magnitude)

hist(quakes$magnitude, main="Quake Magnitude Frequencies", 
     xlab="Magnitude", ylab="Frequency")

boxplot(quakes$magnitude, main = "Quake Magnitudes", 
        xlab= "Magnitude", horizontal=TRUE)
```

## Fit distributions

```{r mag_fit}
# Exponential Distribution

(lambda.exp = 1/mean(quakes$magnitude)) 

# Find Lambda

(loglike.exp <- length(quakes$magnitude)*(log(lambda.exp) - 
                                            lambda.exp*mean(quakes$magnitude))) 

# Log likelihood MLE

(fit.exp <- MASS::fitdistr(quakes$magnitude, "Exponential")) 

# Verify using fitdistr

fit.exp$estimate 

# Extract MLE

fit.exp$loglik 

# Extract Likelihood
```
```{r}
# Gamma Distribution

Meanbar <- mean(quakes$magnitude)

sigmawithhat <- sqrt(sum(quakes$magnitude^2)/length(quakes$magnitude)-Meanbar^2)

lambda.exp <- 1/Meanbar

alpha0 <- Meanbar^2/sigmawithhat^2

lambda0 <- Meanbar/sigmawithhat^2

Meanbar; sigmawithhat; lambda.exp; alpha0; lambda0 

# Print calculations

(fit.gamma <- MASS::fitdistr(quakes$magnitude, "Gamma",
                             start=list(shape=alpha0, rate=lambda0), 
                             lower=0.001))

# Fit Gamma.

fit.gamma$estimate[1]; fit.gamma$estimate[2] 

# Print Gamma Distribution Estimates   
```
```{r}
# Normal Distribution

(fit.normal = MASS::fitdistr(quakes$magnitude, "Normal"))

# Fit Normal Distribution.

fit.normal$estimate[1]; fit.normal$estimate[2]

# Estimates
```

```{r}
# Data histogram + Distribution visualizations

hist(quakes$magnitude, xlab="Magnitude", 
     main=paste("Histogram of Quake Magnitudes (n=", 
                length(quakes$magnitude), ")", sep=""), prob=TRUE) 

# Create a histogram from the quake data sample.

curve(dexp(x, fit.exp$estimate), from=0, 
      to=max(quakes$magnitude), n=1001, col="red", lty=1, add=TRUE) 

# Exp. Dist. Visualization

curve(dgamma(x, fit.gamma$estimate[1], fit.gamma$estimate[2]), from=0, 
      to=max(quakes$magnitude), n=1001, col="green", lty=2, add=TRUE)

# Gamma Distribution Visualization

curve(dnorm(x, fit.normal$estimate[1], fit.normal$estimate[2]), from=0, 
      to=max(quakes$magnitude), n=1001, col="blue", lty=3, add=TRUE) 

# Normal Distribution Visualization.

# Create Legend of each color

lgnd = c(paste("Exp(", round(fit.exp$estimate,3), ")", sep=""),
           paste("Gamma(", round(fit.gamma$estimate[1],3), ",",
                 round(fit.gamma$estimate[2], 3), ")", sep=""),
           paste("Normal(", round(fit.normal$estimate[1], 3), ",",
                 round(fit.normal$estimate[2], 3), ")", sep=""))

legend("right", legend=lgnd, col=c("Red", "Green", "Blue"), 
       lty=c(1, 2, 3))
```

```{r}
# Distribution AICs and ECDF Plot

plot(ecdf(quakes$magnitude), xlab="Magnitude", 
     main=paste("Empirical Distribution of Quake Magnitudes (n=", 
                length(quakes$magnitude), ")", sep=""))

# Add Models

curve(pexp(x, fit.exp$estimate),
      from = 0, to=max(quakes$magnitude), n=1001,
      col = "red", lty = 1, add=TRUE)

curve(pgamma(x, fit.gamma$estimate[1], fit.gamma$estimate[2]),
      from = 0, to=max(quakes$magnitude), n=1001,
      col = "blue", lty = 2, add = TRUE)

curve(pnorm(x, fit.normal$estimate[1], fit.normal$estimate[2]),
      from=0, to=max(quakes$magnitude), n=1001,
      col="green", lty=3, add = TRUE)

# Create Legend

lgnd = c(paste("Exp(", round(fit.exp$estimate,3), ")", sep=""),
           paste("Gamma(", round(fit.gamma$estimate[1],3), ",",
                 round(fit.gamma$estimate[2], 3), ")", sep=""),
           paste("Normal(", round(fit.normal$estimate[1], 3), ",",
                 round(fit.normal$estimate[2], 3), ")", sep=""))

legend("right", legend=lgnd, col=c("Red", "Green", "Blue"), 
       lty=c(1, 2, 3))

# Calculate AIC Values

AIC(fit.normal)

AIC(fit.gamma)

AIC(fit.exp)

```

```{r}
# Exponential Distribution for Time Since Last EQ.

(lambda.exp = 1/mean(quakes$time_since_last_eq)) 

# Find Lambda

(loglike.exp <- length(quakes$time_since_last_eq)*
    (log(lambda.exp) - lambda.exp*mean(quakes$time_since_last_eq))) 

# Log likelihood MLE

(fit.exp <- MASS::fitdistr(quakes$time_since_last_eq, "Exponential")) 

# Verify using fitdistr

fit.exp$estimate 

# Extract MLE

fit.exp$loglik 

# Extract Likelihood
```

```{r}
# Gamma Distribution for Time Since Last EQ.

Xbar <- mean(quakes$time_since_last_eq)

sigmahat <- sqrt(sum(quakes$time_since_last_eq^2)
                 /length(quakes$time_since_last_eq)-Xbar^2)

lambda.exp <- 1/Xbar

alpha0 <- Xbar^2/sigmahat^2

lambda0 <- Xbar/sigmahat^2

Xbar; sigmahat; lambda.exp; alpha0; lambda0 

# Print calculations

(fit.gamma <- MASS::fitdistr(quakes$time_since_last_eq, "Gamma",
                             start=list(shape=alpha0, rate=lambda0), 
                             lower=0.001))

# Fit Gamma.

fit.gamma$estimate[1]; fit.gamma$estimate[2] 

# Print Gamma Distribution Estimates   

```

```{r}
# Normal Distribution for Time Since Last EQ.

(fit.normal = MASS::fitdistr(quakes$time_since_last_eq, "Normal"))

# Fit Normal Distribution.

fit.normal$estimate[1]; fit.normal$estimate[2]

# Estimates

```

```{r}
# Graphic Visualization

# Histogram Plot

hist(quakes$time_since_last_eq, prob=TRUE, xlab="Days",
     main="Time between Earthquakes")

# Add Models

curve(dexp(x, fit.exp$estimate),
      from = 0, to=max(quakes$time_since_last_eq), n=1001,
      col = "red", lty = 1, add=TRUE)

curve(dgamma(x, fit.gamma$estimate[1], fit.gamma$estimate[2]),
      from = 0, to=max(quakes$time_since_last_eq), n=1001,
      col = "blue", lty = 2, add = TRUE)

curve(dnorm(x, fit.normal$estimate[1], fit.normal$estimate[2]),
      from=0, to=max(quakes$time_since_last_eq), n=1001,
      col="green", lty=3, add = TRUE)

# Legend

lgnd = c(paste("Exp(", round(fit.exp$estimate,3), ")", sep=""),
           paste("Gamma(", round(fit.gamma$estimate[1],3), ",",
                 round(fit.gamma$estimate[2], 3), ")", sep=""),
           paste("Normal(", round(fit.normal$estimate[1], 3), ",",
                 round(fit.normal$estimate[2], 3), ")", sep=""))

legend("right", legend=lgnd, col=c("Red", "Green", "Blue"), 
       lty=c(1, 2, 3))

# ECDF Plot

plot(ecdf(quakes$time_since_last_eq), xlab="Time Since Last EQ. (Days)", 
     main=paste("Empirical Distribution of Time Since Last EQ. (n=", 
                length(quakes$time_since_last_eq), ")", sep=""))

# Add Models

curve(pexp(x, fit.exp$estimate),
      from = 0, to=max(quakes$time_since_last_eq), n=1001,
      col = "red", lty = 1, add=TRUE)

curve(pgamma(x, fit.gamma$estimate[1], fit.gamma$estimate[2]),
      from = 0, to=max(quakes$time_since_last_eq), n=1001,
      col = "blue", lty = 2, add = TRUE)

curve(pnorm(x, fit.normal$estimate[1], fit.normal$estimate[2]),
      from=0, to=max(quakes$time_since_last_eq), n=1001,
      col="green", lty=3, add = TRUE)

# Create Legend

lgnd = c(paste("Exp(", round(fit.exp$estimate,3), ")", sep=""),
           paste("Gamma(", round(fit.gamma$estimate[1],3), ",",
                 round(fit.gamma$estimate[2], 3), ")", sep=""),
           paste("Normal(", round(fit.normal$estimate[1], 3), ",",
                 round(fit.normal$estimate[2], 3), ")", sep=""))

legend("right", legend=lgnd, col=c("Red", "Green", "Blue"), 
       lty=c(1, 2, 3))

# Generate QQPlots

qqplot(qexp(ppoints(200), fit.exp$estimate),
       quakes$time_since_last_eq)
abline(0, 1)

qqplot(qgamma(ppoints(200), fit.gamma$estimate[1],
              fit.gamma$estimate[2]),
       quakes$time_since_last_eq)
abline(0, 1)

# Calculate AIC Values

AIC(fit.normal)

AIC(fit.gamma)

AIC(fit.exp)

```



End.

\newpage
\appendix
# Appendix: R Environment
```{r r_environ}
sessionInfo()
```